---
title: Static visualizations
format: 
    html:
        code-fold: true
---

```{r}
#| include: false
#| label: setup_r
library(tidyverse)
library(rio)
library(plotly)
library(reticulate)
library(ggsci)
library(showtext)
font_add_google("Atkinson Hyperlegible")
showtext_auto()
reticulate::use_virtualenv('/Users/abhijit/.pyenv/versions/3.12.7/envs/dsan5200-3.12.7', required=TRUE)
theme_set(
    theme_bw() + 
        theme(text = element_text(family = "Atkinson Hyperlegible", color = "#145da0", face = "bold"))
)
```

```{python}
#| label: setup_py
#| include: false
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import seaborn.objects as so
```
## The grammar of graphics

We're familiar with the [grammar of graphics](https://vita.had.co.nz/papers/layered-grammar.html) from our familiarity with `ggplot2`. These principles have also been applied to develop the semantics of several Javascript plotting libraries, including [Vega-Lite](https://vega.github.io/vega-lite/) and [Observable Plot](https://observablehq.com/plot/). Let's see how the grammar is implemented. 

We'll start with some data from the US Energy Information Administration, looking at energy consumption over time of different energy sources. 

```{r}
#| cache: false
energy <- read_csv('data/us_energy.csv')
ojs_define(energy_long = energy)
# head(energy)
```



```{ojs}
energy = transpose(energy_long).map(
    item => {
        item.year = d3.utcFormat("%Y")(new Date(item.year, 0,1));// convert to date 
        return item;
    }
)
Inputs.table(energy,{
    header: {
        year: "Year",
        type: "Source",
        quadrillion_btu : "Energy use (Quadrillion BTU)"
    }
})
```

Let's start with some familiar visualizations:

```{r}
#| echo: true
ggplot(energy, aes(x = year, y = quadrillion_btu, color = type)) +
    geom_point() + geom_line()+
    scale_color_npg() +
    labs(x = "Year", y = "Energy (Quadrillion BTU)", color = "Source",
    title="Energy use by source, 1973-2022")
```


```{python}
#| eval: false
energy = pd.read_csv('data/us_energy.csv')
(so.Plot(energy, x = "year", y = "quadrillion_btu", color = "type")
    .add(so.Dot())
    .add(so.Lines())
    )
```

::: {.panel-tabset}

## Observable Plot
```{ojs}
//| echo: true

Plot.plot({
    title: "Energy use by source, 1973-2022",
    x: { label: "Year", type:'time'},
    y: { label: "Energy (Quadrillion BTU)"},
    color: {scheme: "Tableau10", legend: true, label: "Source", columns: 1,
        style: {float: "right"}
    },
    stroke: {scheme: "Tableau10", legend: true, label: "Source", columns: 5},
    style: {
        backgroundColor: "white",
        color: "#145da0",
        fontFamily: "Atkinson Hyperlegible",
        fontSize: "14px",
        padding: "16px"
    },
    marks: [
        Plot.dot(energy, 
            {x: 'year',
            y: "quadrillion_btu", 
            fill: "type",
            tip: true,
            }),
        Plot.line(energy, 
            {x: 'year',
            y: "quadrillion_btu", 
            stroke: "type"}),
        Plot.frame(), // add box around plot
    ]
})
```

::: {.callout-tip}
Plot accepts many pre-specified color scales. See [here](https://observablehq.com/plot/features/scales#color-scales). Legend specs can be found [here](https://observablehq.com/plot/features/legends)
:::

## Vega-Lite
```{ojs}
//| echo: true

{
    const points = vl.markPoint()
    .data(energy)
    .encode(
        vl.x().fieldT('year').axis({title: "Year"}),
        vl.y().fieldQ('quadrillion_btu').axis({title: "Energy use (Quadrillion BTU)"}),
        vl.color().fieldN('type').title("Source"),
        vl.tooltip(['year','type','quadrillion_btu'])
    ).width(400).height(400);
const lines = vl.markLine()
    .data(energy)
    .encode(
        vl.x().fieldT('year').axis({title: "Year"}),
        vl.y().fieldQ('quadrillion_btu').axis({title: "Energy use (Quadrillion BTU)"}),
        vl.color().fieldN('type')
    ).width(400).height(400);
return vl.layer(lines, points).title("Energy use by source, 1973-2022").render();
}
```

## Plotly (R)


```{r}
#| echo: true
plot_ly(energy, 
    x = ~year,
    y = ~quadrillion_btu,
    color = ~type,
    text = ~type,
    type = 'scatter',
    mode = 'lines+markers',
        hovertemplate = paste(
            "<b>Year:</b> %{x}",
            "<b>Energy use:</b> %{y:.2f}",
            "<b>Source:</b> %{text}",
            sep = "\n"
        )
    ) |> 
    layout(
        title = "Energy use by source, 1973-2022",
        xaxis=list(title='Year'),
        yaxis=list(title = "Energy (Quadrillion BTU)",
        font = list(
            family="Gravitas One",
            color = "#145da0"
        ),
        legend = list(
            font = list(
                family = "Gravitas One",
                size = 14,
                color = "#145da0")))
    )
```

:::